---
title: 'Reproducible Research: Peer Assessment 1'
output:
  html_document:
    keep_md: yes
  pdf_document: default
  word_document: default
---


## Loading and preprocessing the data
The activity.csv was created by extracting the contents of the zip file.  It was read simply by read.csv() and a summary statistics are displayed
``` {r load_data, echo="true"}
data = read.csv('activity.csv')
summary(data)
```

## What is mean total number of steps taken per day?
``` {r mean_values}
stepsByDay <-  aggregate(data$steps, by=list(data$date),FUN=sum,na.rm=TRUE)
hist(stepsByDay$x,main="Histogram of total steps per day",  xlab="number of steps per day")
mean(stepsByDay$x)
median(stepsByDay$x)
```


## What is the average daily activity pattern?
``` {r average_daily}
stepsByInterval <-  aggregate(data$steps, by=list(data$interval),FUN=mean,na.rm=TRUE)
times<-strptime(formatC(stepsByInterval$Group.1,width=4,format="d", flag="0"),'%H%M')
plot(times,stepsByInterval$x, type='l', main="Daily activity pattern",  ylab="number of steps per interval", xlab="Time of day")
maxTime <- subset(stepsByInterval, stepsByInterval$x==max(stepsByInterval$x))
format(strptime(formatC(maxTime$Group.1,width=4,format="d", flag="0"),'%H%M'),'%H:%M')
```
## Imputing missing values
The strategy to impute missing values is to extract the average value for the specific interval using a function to compute the lookup.

``` {r imputed_values}
getMean <- function(tst) {
  tst2=tst[3]
  if (!is.nan(as.numeric(tst2))) 
    tst[1] <- (stepsByInterval$x[as.numeric(tst2)==stepsByInterval$Group.1]);
}
imputed<-data
imputedValues<- apply(imputed,1,function(x) getMean(x))
imputed$imputedSteps <- imputedValues
imputed <- transform(imputed,steps=ifelse(is.na(steps),imputedSteps,steps))

stepsByDayImputed <-  aggregate(imputed$steps, by=list(imputed$date),FUN=sum,na.rm=TRUE)
hist(stepsByDayImputed$x,main="Histogram with missing values imputed",  xlab="number of steps per day")
mean(stepsByDayImputed$x)
median(stepsByDayImputed$x)
```
The effect imputing the values is to push the distribution of the number of steps to a more normal distribution
## Are there differences in activity patterns between weekdays and weekends?
``` {r weekend_difference}

imputed$weekend <- factor(chron::is.weekend(imputed$date),labels=c("weekday","weekend"))
groupSteps <- aggregate(imputed$steps, by =list(interval=imputed$interval,weekend=imputed$weekend),FUN=mean)
library(ggplot2)
g <- ggplot(groupSteps, aes(interval, x))
g + geom_line() + facet_grid(weekend ~ .) +
labs(title="Daily activity pattern by weekday type") +
labs(y="average number of steps") +
labs(x="time of day") 

```
This graph shows that during a weekday, activities peak during the morning, wheras during the weekend activity is more evenly distributed throughout the day